Co-methylation review

### 2015-04-20 Rmarkdown/biology_background.Rmd
Replace ‘nucleotide’ with ‘bp’ where appropriate.

[^snvs_snps]: An individual nucleotide that is variable in the population is called a single nucletoide variant (_SNV_). A SNV that is frequently variable in the population, say at least $1\%$ of the population are variable at that site, is called a single nucleotide polymorphism (SNP). All SNPs are SNVs but the converse is not true.

>>>>

[^snvs_snps]: A base position that is variable in the population is called a single nucleotide variant (_SNV_). A SNV that is frequently variable in the population, say at least $1\%$ frequency, is called a single nucleotide polymorphism (SNP). All SNPs are SNVs but the converse is not true.

### 2015-04-20 Rmarkdown/comethylation_review.Rmd
Replace ‘nucleotide’ with ‘bp’ where appropriate.

[^CHH_only]: The $10$ nucleotide period was later observed for CG and CHG methylation as well, at least for nucleosome bound DNA \citep{Chodavarapu:2010iqa}

@@

An $8$ to $10$ bp period in the fitted cubic spline is observed in most, but not all, of these nine plots. This periodicity is very weak in several contexts and is absent for exonic CpGs. The weakness of the signal is partially attributable to the small number of "methylcytosines" in the CHG and CHH contexts in human samples. More importantly, the reported $8$ to $10$ bp periods are weaker than a $3$ bp period for "methylcytosines" at exonic CpGs. Unlike the $8$ to $10$ nucleotide period, the $3$ nucleotide period can clearly be observed in the raw counts, although the cubic spline smoother seemingly ignores it. \citeauthor{Lister:2009hy} do not comment on the $3$ bp period.

@@

In fact, there are well known periodicities in the distribution of nucleotides in genomes. For example, in eukaryotic genomes there is a $10$ bp period of G+C dinucleotides that acts out of phase with a $10$ bp period of A+T dinucleotides for nucleosome-bound DNA \citep{Segal:2006iz} and a "well-known period-3 oscillation in coding regions" of the genome \citep{Li:1997ts}; the former likely explains the $8$ to $10$ nucleotide period and the latter the $3$ bp period for exonic CpGs.

@@

\citet{Chodavarapu:2010iqa} use MNase-seq to study nucleosome positioning and the relationship to DNA methylation, including co-methylation, in _Arabidopsis_ and an embryonic human stem cell line (_HSF1_). The authors plot a "weighted average methylation", which I presume to be a weighted version of the $\beta$-values, against the distance from predicted nucleosome start sites, which are obtained via MNase-seq. The nucleosome start sites anchor the methylation observations to a common grid and this plot is used to show that nucleosome bound DNA has a greater level of methylation than DNA that is not bound to nucleosomes. A $10$ bp periodicity is readily observed in these plots, consistent with the discovery of $10$ nucleotide periodicities in the autocorrelation of $\beta$-values published by \cite{Cokus:2008fc}. Again, the observed periodicity is confirmed by analysing a Fourier transform of this correlation signal.

@@

To summarise, current technologies mean that we are limited to studying within-fragment co-methylation at loci that are within regions of approximately $200-250$ nucleotides.

@@

The sequencing technology at the time of this publication produced very short reads, on average only $31$ bp of usable sequence. For the within-fragment co-methylation results, \citeauthor{Cokus:2008fc} first identify all reads containing multiple CHH loci. Then, for a pair of CHH loci in the same read, they estimate the probability that the second CHH in the pair is methylated given that the first CHH in the pair is methylated. This is estimated by simply counting the number of times this occurs and dividing by the number of pairs of CHH loci that occur within the same read[^cokus_clarification]. They estimate this probability separately for each distance between CHH loci and separately for each of the five _Arabidopsis_ chromosomes. \citeauthor{Cokus:2008fc} then plot the average across the five chromosomes as a function of the number of nucleotides between the CHH loci. The plot includes pointwise $95\%$ confidence intervals, computed under an assumption of normality, and a running average of these averages. This plot, reproduced in Figure \ref{fig:Cokus_2008_Sup_Fig_14}, shows evidence of an approximately $10$ bp period, supporting the interpretation of the results observed in the autocorrelation of the $\beta$-values.

@@

__UP TO HERE: Work through the logic of this next paragraph__

@@

While it has long been observed that DNA methylation at nearby loci is highly correlated, pinning down exactly what this means, and its implications, is surprisingly difficult. As I have shown, there are problems with most, if not all, the previous analyses, ranging from a lack of methodological detail and limited data, through to serious flaws in the methodology or its implementation. Moreover, despite numerous papers including an analysis of co-methylation, none of these have includes software to reproduce their method, further complicating the assessment of these methods.

>>>>

[^CHH_only]: The $10$ bp period was later observed for CG and CHG methylation as well, at least for nucleosome bound DNA \citep{Chodavarapu:2010iqa}

@@

An $8$ to $10$ bp period in the fitted cubic spline is observed in most, but not all, of these nine plots. This periodicity is very weak in several contexts and is absent for exonic CpGs. The weakness of the signal is partially attributable to the small number of "methylcytosines" in the CHG and CHH contexts in human samples. More importantly, the reported $8$ to $10$ bp periods are weaker than a $3$ bp period for "methylcytosines" at exonic CpGs. Unlike the $8$ to $10$ bp period, the $3$ bp period can clearly be observed in the raw counts, although the cubic spline smoother seemingly ignores it. \citeauthor{Lister:2009hy} do not comment on the $3$ bp period.

@@

In fact, there are well known periodicities in the distribution of nucleotides in genomes. For example, in eukaryotic genomes there is a $10$ bp period of G+C dinucleotides that acts out of phase with a $10$ bp period of A+T dinucleotides for nucleosome-bound DNA \citep{Segal:2006iz} and a "well-known period-3 oscillation in coding regions" of the genome \citep{Li:1997ts}; the former likely explains the $8$ to $10$ bp period and the latter the $3$ bp period for exonic CpGs.

@@

\citet{Chodavarapu:2010iqa} use MNase-seq to study nucleosome positioning and the relationship to DNA methylation, including co-methylation, in _Arabidopsis_ and an embryonic human stem cell line (_HSF1_). The authors plot a "weighted average methylation", which I presume to be a weighted version of the $\beta$-values, against the distance from predicted nucleosome start sites, which are obtained via MNase-seq. The nucleosome start sites anchor the methylation observations to a common grid and this plot is used to show that nucleosome bound DNA has a greater level of methylation than DNA that is not bound to nucleosomes. A $10$ bp period is readily observed in these plots, consistent with the discovery of $10$ bp period in the autocorrelation of $\beta$-values published by \cite{Cokus:2008fc}. Again, the observed periodicity is confirmed by analysing a Fourier transform of this correlation signal.

@@

To summarise, current technologies mean that we are limited to studying within-fragment co-methylation at loci that are within regions of approximately $200-250$ bp.

@@

The sequencing technology at the time of this publication produced very short reads, on average only $31$ bp of usable sequence. For the within-fragment co-methylation results, \citeauthor{Cokus:2008fc} first identify all reads containing multiple CHH loci. Then, for a pair of CHH loci in the same read, they estimate the probability that the second CHH in the pair is methylated given that the first CHH in the pair is methylated. This is estimated by simply counting the number of times this occurs and dividing by the number of pairs of CHH loci that occur within the same read[^cokus_clarification]. They estimate this probability separately for each distance between CHH loci and separately for each of the five _Arabidopsis_ chromosomes. \citeauthor{Cokus:2008fc} then plot the average across the five chromosomes as a function of the number of bp between the CHH loci. The plot includes pointwise $95\%$ confidence intervals, computed under an assumption of normality, and a running average of these averages. This plot, reproduced in Figure \ref{fig:Cokus_2008_Sup_Fig_14}, shows evidence of an approximately $10$ bp period, supporting the interpretation of the results observed in the autocorrelation of the $\beta$-values.

@@

@@

While it has long been observed that DNA methylation at nearby loci is highly correlated, pinning down exactly what this means, and its implications, is surprisingly difficult. As I have shown, there are problems with most, if not all, previous analyses. These problems range from a lack of methodological detail and limited data, through to serious flaws in the methodology or its implementation. Moreover, despite numerous papers including an analysis of co-methylation, none of these have includes software to reproduce their method, further complicating the assessment of these methods.
